Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-09-14 RNARedPrint.tex
Updates from ShareLaTeX

\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}

\newtheorem{observation}[theorem]{Observation}

@@

\begin{abstractbox}
\begin{abstract}
  \parttitle{Background}
  The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches profit strongly from effective in-silico methods, which impact their cost and feasibility. \\
%
  \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
  design methods: the sampling of sequences for multi-target RNA design. For this task, we present the efficient (fixed-parameter tractable) approach \ourprog{}.
  % 
  Modeling
  the problem by specializing constraint networks, \ourprog{} supports generic
  Boltzmann-weighted sampling, which enables the positive design of RNA sequences
  with specific free energies (for each of multiple, possibly pseudoknotted, target structures) and \GCb-content.
%
  Securing the theoretical foundations of our method, we establish the $\#${\sf P}-hardness of uniform sampling. 
%
  Finally, we empirically study general properties of the approach and generate
  biologically relevant multi-target Boltzmann-weighted designs for a
  common design benchmark, including pseudoknots.
  Generating seed sequences with \ourprog{}, we demonstrate significant improvements over the previously best multi-target sampling strategy.\\
%
  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible and---as we demonstrate--effective approach to multi-target RNA design, which promises broad applicability and extensibility.

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
\parHead{Design, applications and motivation for multiple design.}Synthetic biology strives for the engineering of artificial biological
systems, promising broad applications in biology, biotechnology and
medicine. Centrally, this requires the design of biological
macromolecules with highly specific properties and programmable functions.
RNAs are particularly well-suited tools for
rational design targeting specific functions~\citep{Kushwaha2016}: on the one hand, RNA function is tightly
coupled to the formation of secondary structure, as well as changes in
base pairing propensities and the accessibility of regions, e.g.\ by
burying or exposing interaction sites~\citep{Rodrigo2014}; on the other hand, the
thermodynamics of RNA secondary structure is well understood and its prediction is
computationally tractable~\citep{McCaskill1990}. Thus,  structure can serve as effective
proxy for rational design approaches, ultimately targeting catalytic or regulatory functions~\citep{Zhang2013}.

@@

\parHead{Motivation.}
While uniform sampling in \Software{RNAdesign}/\RNAblueprint{} turned out to be very successful,
the work on these tools gave no definitive answer on the \emph{complexity of uniform sampling for multi-target design}. While time and space requirements grow exponential in the general case,
the decomposition strategy seems to allow practical applications.
Better understanding of the true complexity bounds is still highly relevant for the application of such methods, in particular for applications in more expressive settings (e.g. Boltzmann sampling as discussed in this work or sampling satisfying multiple additional constraints and properties). All such extensions are still bounded by the complexity of the multi-target uniform sampling problem, where counting of valid designs is the dominating step (note that stochastic backtracking can be performed in linear time per sample).

@@

To answer these questions, we introduce a generic framework (illustrated in~Fig.~\ref{fig:workflow}) enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures. As main algorithmic contribution of this work, we devise a dynamic programming algorithm, based on a tree decomposition, to compute partition functions and sample sequences from the Boltzmann distribution%
% (Subsection~\ref{sec:PF})
. We show that these algorithms are time and space-efficient depending on a specific property of the tree decomposition, namely the treewidth; in short, our algorithms are fixed-parameter tractable (FPT).%
% Subsection~\ref{sec:complexity}
%
Due to the generality of our method, we can moreover strongly limit this parameter in practice by using state-of-the-art tree decomposition algorithms.
%Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
By evaluating (partial) sequences in a specialized weighted constraint
network, we support arbitrary multi-ary constraints and thus
arbitrarily complex energy models,
notably subsuming all commonly
used RNA energy models%
%(Subsection~\ref{sec:energy_models})
.  Moreover, we describe an \Def{adaptive
  sampling} strategy to control the free energies of the individual
target structures and \GCb\%%
% (SubSection~\ref{sec:multiBoltzmann})
. %

@@

Eventually this strategy allows generating biologically relevant
multi-target designs, as we demonstrate in our application 
to a large set of multi-target RNA design instances from a
meaningful, representative benchmark.

\section*{Methods}
\subsection*{Problem statement}
\label{sec:problem-statement}
The main computational problem addressed in this work is to generate sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$ that achieving prescribed free-energies and $\GCb\%$.
Despite their apparent diversity, such criteria can be homogeneously abstracted as constraints on the value of \Def{features}, each defined as sums over atomic feature contributions depending on subsets of sequence positions.

\paragraph{Central computational problem.}
Let us consider $k$ features $F_1,\dots F_k$ to target all of them simultaneously. We base our strategy for targeting the $k$ features $F_1,\dots,F_k$ on generating samples from a Boltzmann distribution parameterized by weights $\pi_1,\dots,\pi_k$, one for each feature. These weights essentially allows us to shift the sample distributions towards specific targets.
Sampling from this distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight:
\begin{equation}
\label{eq:sample-distribution}
\Pr(S|\pi_1,\dots,\pi_k) \propto \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
\end{equation}

The workhorse of our approach is the efficient (more precisely, fixed-parameter tractable) computation of feature-dependent partition functions over sequences.
Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\% all in terms of simple atomic energy contributions.
For weights $\pi_1,\dots,\pi_k$, one wants to compute the partition function
  \begin{equation}
    \label{eq:mainproblem}
    \partfun{\pi_1,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
  \end{equation}

\paragraph{Representing \GCb\%-content and energies as features.}
Formally, we define a \Def{feature $F$} as function on sequences, whose value is obtained by summing over an associated set of \Def{atomic feature contributions} $f_1, f_2\ldots$, briefly called \emph{contributions}, that each are associated with a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
$$
F(S) =\!\!\!\!\!\!\sum_{\substack{\text{contribution $f_i$ of $F$}\\\text{with $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\! f_i(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
$$

As simple example, the $\GCb\%$ feature is expressed by $n$ contributions $f^{\GCb}_i$ where $f^{\GCb}_i(\{i\mapsto S_i\})$ equals $-1$ if $S_i=\Gb$ or $\Cb$, and $0$ otherwise.

@@

In a simple \Def{base pair energy model}, the free-energy of a target structure $R$ is expressed as feature with contributions $f^R_{ij}(\{i\mapsto x, j\mapsto y\})=\EbpSym(x,y)$ for each $(i,j)\in R,$ such that $F^R(S)=\EbpSym(S,R)$.

Note that for \GCb\% as well as for the base pair energy features, the contributions depend on the nucleotides at only a small, constantly bounded number of sequence positions (one and two positions, respectively). This is important for the feasibility of our approach, which is designed to profit from contributions that each depend on only few sequence positions. 
We emphasize that nevertheless, the same framework can capture more expressive RNA energy models like the Turner energy model~\citep{Turner2009}, since such models generally define a function $E$ that maps a pair of a sequence $S$ and a structure $R$ to a real value, the \Def{energy} $E(S,R)$. Here, our approach automatically adapts to the more complex energy functions, leading to a controlled increase of the computational complexity.

@@

To formalize our algorithms, which iteratively merges evaluations of partial solutions, we extend the idea of atomic feature contributions, which are evaluated at sets of the form $\{x_1\mapsto v_1, \dots, x_p\mapsto v_d\}$. Let us call the latter object \Def{partial sequence}; they are handy to specify partial knowledge on the entire sequence at some point of the algorithm. Easily, we can extend the definition of contributions $f$ to sets $\{x_1\mapsto v_1, \dots, x_p\mapsto v_d\}$, where $\{x_1\dots x_d\}$ is any super-set of $\dep(f)$ by ignoring the superfluous assignments $x\mapsto v$, where $x\not\in\dep(f)$.

@@

\newcommand{\weighttransform}[2]{{#1}_{||{#2}}}

For algorithmic treatment, it is convenient to encode the weight $\pi$ of each feature in its contributions, such that all contributions can be handled uniformly. For this purpose, we define the \Def{transformed contribution} 
$\weighttransform{f}{\pi}(S) := \ln(\pi)f(S),$
such that $exp(-\weighttransform{f}{\pi}(S))=\pi^{-f(S)}.$ 
%

Let us now specify the concrete set $\F$ of contributions that we use
for the design in the base pair model targeting \GCb\% and structures $\R$ with weights $\pi_0,\dots,\pi_k$. The set $\F$ thus consists of
\begin{itemize}
\item the transformed contributions $\weighttransform{f^{\GCb}_i}{\pi_0}$ for the ~$\GCb\%$ feature ($i=1,\dots,n$)
\item the transformed contributions $\weighttransform{f^{R_\ell}_{ij}}{\pi_\ell}$for each structure $R_\ell\in\R$ and $(i,j)\in R_\ell.$
\end{itemize}

@@

We compute the partition function (as specified by $\F$) by dynamic programming based
on a tree decomposition of the $\F$'s dependency graph.
Note that analogous algorithms could be constructed to count valid sequences or minimize energy.
%This yields fixed parameter tractable (FPT) algorithms, based on the treewidth.

>>>>

\newtheorem{theorem}{Theorem}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{observation}[theorem]{Observation}

@@

\begin{abstractbox}
\begin{abstract}
  \parttitle{Background}
  The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches profit strongly from effective in-silico methods, which impact their cost and feasibility. \\
%
  \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
  design methods: the generation of sequences with multiple structural targets. For this task, we present the efficient (fixed-parameter tractable) approach \ourprog{}.
  % 
  Modeling
  the problem by specializing constraint networks, \ourprog{} supports generic
  Boltzmann-weighted sampling, which enables the positive design of RNA sequences
  with specific free energies (for each of multiple, possibly pseudoknotted, target structures) and \GCb-content.
%
  Securing the theoretical foundations of our method, we establish the $\#${\sf P}-hardness of uniform sampling. 
%
  Finally, we empirically study general properties of the approach and generate
  biologically relevant multi-target Boltzmann-weighted designs for a
  common design benchmark, including pseudoknots.
  Generating seed sequences with \ourprog{}, we demonstrate significant improvements over the previously best multi-target sampling strategy.\\
%
  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible and---as we demonstrate--effective approach to multi-target RNA design, which promises broad applicability and extensibility.

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
\parHead{Design, applications and motivation for multiple design.}Synthetic biology strives for the engineering of artificial biological
systems, promising broad applications in biology, biotechnology and
medicine. Centrally, this requires the design of biological
macromolecules with highly specific properties and programmable functions.
RNAs are particularly well-suited tools for
rational design targeting specific functions~\citep{Kushwaha2016}: on the one hand, RNA function is tightly
coupled to the formation of secondary structure, as well as changes in
base pairing propensities and the accessibility of regions, e.g.\ by
burying or exposing interaction sites~\citep{Rodrigo2014}; on the other hand, the
thermodynamics of RNA secondary structure is well understood and its prediction is
computationally tractable~\citep{McCaskill1990}. Thus, in rational design approaches, structure can serve as effective
proxy for, the ultimately targeted, catalytic or regulatory functions~\citep{Zhang2013}.

@@

\parHead{Motivation.}
While uniform sampling in \Software{RNAdesign}/\RNAblueprint{} turned out to be highly beneficial for general RNA design,
the work on these tools gave no definitive answer on the \emph{complexity of uniform sampling for multi-target design}. While time and space requirements grow exponential in the general case,
the decomposition strategy seems to allow practical applications.
Better understanding of the true complexity bounds is still highly relevant for the application of such methods, in particular for applications in more expressive settings (e.g. Boltzmann sampling as discussed in this work or sampling satisfying multiple additional constraints and properties). All such extensions are still bounded by the complexity of the multi-target uniform sampling problem, where counting of valid designs is the dominating step (note that stochastic backtracking can be performed in linear time per sample).

@@

To answer these questions, we introduce a generic framework (illustrated in~Fig.~\ref{fig:workflow}) enabling efficient Boltzmann-weighted sampling over RNA sequences with multiple target structures. As main algorithmic contribution of this work, we devise a dynamic programming algorithm, based on a tree decomposition, to compute partition functions and sample sequences from the Boltzmann distribution%
% (Subsection~\ref{sec:PF})
. We show that these algorithms are time and space-efficient depending on a specific property of the tree decomposition, namely the treewidth; in short, our algorithms are fixed-parameter tractable (FPT).
% Subsection~\ref{sec:complexity}
%
Due to the generality of our method, we can moreover strongly limit this parameter in practice by using state-of-the-art tree decomposition algorithms.
%Uniform sampling is handled as a special case of Boltzmann sampling, where each valid sequence receives energy zero and---consequently---computing partition functions specializes to counting.
By evaluating (partial) sequences in a specialized weighted constraint
network, we support arbitrary multi-ary constraints and thus
arbitrarily complex energy models,
notably subsuming all commonly
used RNA energy models%
%(Subsection~\ref{sec:energy_models})
.  Moreover, we describe an \Def{adaptive
  sampling} strategy to control the free energies of the individual
target structures and \GCb\%%
% (SubSection~\ref{sec:multiBoltzmann})
. %

@@

Eventually this strategy allows generating biologically relevant
multi-target designs, as we demonstrate by performing negative multi-target RNA design for a comprehensive, 
meaningful benchmark set~\citep{Taneda2015}.

\section*{Methods}
\subsection*{Problem statement}
\label{sec:problem-statement}
The main computational problem addressed in this work is to generate sequences over the alphabet $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$ with prescribed energies, for each single target structure, and $\GCb\%$.
Remarkably, such criteria can be homogeneously abstracted as constraints on the value of \Def{features}, each defined as sums over atomic feature contributions depending on limited subsets of sequence positions.

\paragraph{Central computational problem.}
Let us consider $k$ features $F_1,\dots F_k$ to target all of them simultaneously. We base our strategy for targeting the $k$ features $F_1,\dots,F_k$ on generating samples from a Boltzmann distribution parameterized by weights $\pi_1,\dots,\pi_k$, one for each feature. These weights essentially allow us to shift the sample distributions towards specific targets.
Sampling from this distribution requires to compute corresponding partition functions, such that we can draw sequences with probabilities proportional to their Boltzmann weight:
\begin{equation}
\label{eq:sample-distribution}
\Pr(S|\pi_1,\dots,\pi_k) \propto \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
\end{equation}

The workhorse of our approach is the efficient (more precisely, fixed-parameter tractable) computation of feature-dependent partition functions over sequences.
Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\%, all in terms of simple atomic energy contributions.
For weights $\pi_1,\dots,\pi_k$, one wants to compute the partition function
  \begin{equation}
    \label{eq:mainproblem}
    \partfun{\pi_1,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
  \end{equation}

\paragraph{Representing \GCb\%-content and energies as features.}
Formally, we define a \Def{feature $F$} as function on sequences, whose value is obtained by summing over an associated set of \Def{atomic feature contributions}, briefly called \emph{contributions}. Each feature $f$ is associated with a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
$$
F(S) =\!\!\!\!\!\!\sum_{\substack{\text{$f$: $f$ is contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\! f(\{x_1\mapsto S_{x_1},\ldots,x_p\mapsto S_{x_p}\}).
$$

Thus, the $\GCb\%$ feature is expressed by $n$ contributions $f^{\GCb}_i$ where $f^{\GCb}_i(\{i\mapsto S_i\})$ equals $-1$ if $S_i=\Gb$ or $\Cb$, and $0$ otherwise $(i=1,\dots,n)$.

@@

In a simple \Def{base pair energy model}, the energy of a target structure $R$ is expressed as feature with contributions $f^R_{ij}(\{i\mapsto x, j\mapsto y\})=\EbpSym(x,y)$ for each $(i,j)\in R,$ such that $F^R(S)=\EbpSym(S,R)$.

Note that for \GCb\% as well as for the base pair energy features, the contributions depend on the nucleotides at only a small, constantly bounded number of sequence positions (one and two positions, respectively). This is important for the efficiency of our approach, which is designed to profit from contributions that each depend on only few sequence positions. 
We emphasize that nevertheless, the same framework can capture more expressive RNA energy models like the Turner energy model~\citep{Turner2009}, since such models generally define a function $E$ that maps a pair of a sequence $S$ and a structure $R$ to a real value, the \Def{energy} $E(S,R)$. Here, our approach automatically adapts to the more complex energy functions, leading to a controlled increase of the computational complexity.

@@

To formalize our algorithms, which iteratively merge evaluations of partial solutions, we extend the idea of atomic feature contributions, which are evaluated at sets of the form $\{x_1\mapsto v_1, \dots, x_d\mapsto v_p\}$. Let us call the latter object \Def{partial sequence}; they are handy to specify partial knowledge on the entire sequence at some point of the algorithm. Easily, we can extend the definition of contributions $f$ to sets $\{x_1\mapsto v_1, \dots, x_p\mapsto v_p\}$, where $\{x_1\dots x_p\}$ is any super-set of $\dep(f)$ by ignoring the superfluous assignments $x\mapsto v$, where $x\not\in\dep(f)$.

@@

%\newcommand{\weighttransform}[2]{{#1}_{||{#2}}}

For algorithmic treatment, it is convenient to encode the weight $\pi$ of each feature in its contributions, such that all contributions can be handled uniformly. This transformation works by multiplying all contributions with $ln(\pi)$, where $\pi$ is the weight of the corresponding feature, since then
$exp(-\ln(\pi)f(S))=\pi^{-f(S)}.$

%For this purpose, we define the \Def{transformed contribution} 
%$\weighttransform{f}{\pi}(S) := \ln(\pi)f(S),$
%such that $exp(-\weighttransform{f}{\pi}(S))=\pi^{-f(S)}.$ 
%
Let us now specify the concrete set $\F$ of contributions that we use
for the design in the base pair model targeting \GCb\% and structures $\R$ with weights $\pi_0,\dots,\pi_k$. The set $\F$ thus consists of
\begin{itemize}
\item the transformed contributions $\ln(\pi_0)f^{\GCb}_i$ for the ~$\GCb\%$ feature ($i=1,\dots,n$)
\item the transformed contributions $\ln(\pi_\ell)f^{R_\ell}_{ij}$for each structure $R_\ell\in\R$ and $(i,j)\in R_\ell.$
\end{itemize}

@@

We compute the partition function (as specified by $\F$) by dynamic programming based
on a tree decomposition of $\F$'s dependency graph.
Note that analogous algorithms could be constructed to count valid sequences or minimize energy.
%This yields fixed parameter tractable (FPT) algorithms, based on the treewidth.
